---
title: "class09"
author: Alishba Qadri
format: pdf
editor: visual
---

# 2 Importing candy data

```{r}
#load dataset
candy_file <- "https://raw.githubusercontent.com/fivethirtyeight/data/master/candy-power-ranking/candy-data.csv"

#inspect data
candy = read.csv(candy_file, row.names=1)
head(candy)
```

> **Q1. How many different candy types are in this dataset?**

> **A1. There are 85 different candy types in this dataset.**

```{r}
dim(candy)
```

> **Q2. How many fruity candy types are in the dataset?**

> **A2. There are 38 fruity candy types in the dataset.**

```{r}
sum(candy$fruity)
```

```{r}
# find the winpercent value for Twix
candy["Twix", ]$winpercent
```

```{r}
#alternate way to find the winpercent value using dplyr
library(dplyr)

candy |> 
  filter(row.names(candy)=="Twix") |> 
  select(winpercent)
```

> **Q3. What is your favorite candy (other than Twix) in the dataset and what is it’s winpercent value?**

> **A3. My favorite candy is Almond Joy. Its winpercent value is 50.34755.** 

```{r}
# find the winpercent value of Almond Joy
candy["Almond Joy",]$winpercent
```

> **Q4. What is the winpercent value for “Kit Kat”?**

> **A4. The winpercent value is 76.7686.**

```{r}
# find  the winpercent valye of Kit Kat
candy["Kit Kat",]$winpercent
```

> **Q5. What is the winpercent value for “Tootsie Roll Snack Bars”?**

> **A5. The winpercent value is 49.6535.**

```{r}
# find the winpercent value of Tootsie Rolls
candy["Tootsie Roll Snack Bar",]$winpercent
```

```{r}
# load skimr package
library("skimr")
# look at skimr dataset overview
skim(candy)
```

> **Q6. Is there any variable/column that looks to be on a different scale to the majority of the other columns in the dataset?**

> **A6. Yes, winpercent is the only variable that contains  percentages (on a scale from 0 to 100) rather than decimals (on a scale from 0 to 1).** 

> **Q7. What do you think a zero and one represent for the candy$chocolate column?**

> **A7. A zero represents that the candy does not contain chocolate, while a one represents that the candy does contain chocolate.** 

```{r}
candy$chocolate
```

# 3 Exploratory analysis

> **Q8.Plot a histogram of winpercent values using both base R an ggplot2.**

```{r}
# create histogram of winpercent values using base R
hist(candy$winpercent)
```

```{r}
# load ggplot2
library(ggplot2)

# create histogram of winpercent values using ggplot
ggplot(candy) +
  aes(winpercent) +
  geom_histogram(bins = 10, fill="grey", color = "black") 
```

> **Q9. Is the distribution of winpercent values symmetrical?**

> **A9. No, the distribution is not symmetrical. It seems to be skewed right, with the maximum frequency winpercent values being at the 40-50 bar. **

> **Q10. Is the center of the distribution above or below 50%?**

> **A10. No, the center of distribution (median) is 47.83%, thereby indicating it is below 50%.** 

```{r}
# observe distribution of data
summary(candy$winpercent)
```

> **Q11. On average is chocolate candy higher or lower ranked than fruit candy?**

> **A11. On average, chocolate candy is higher ranked than fruity candy, with a mean winpercent value of 60.92%. Fruity candy has a mean value of 44.12%.** 

```{r}
# t-test of chocolate
chocolate <- candy$winpercent[as.logical(candy$chocolate)]
t.test(chocolate)

#t-test of fruity candy
fruit <- candy$winpercent[as.logical(candy$fruity)]
t.test(fruit)
```

> **Q12. Is this difference statistically significant?**

> **A12. Yes, this difference is statistically significant, as the data has a p-value < 0.05. Therefore, it is extremely unlikely that the results are based on random chance.**

# 4. Overall Candy Rankings

> **Q13. What are the five least liked candy types in this set?**

> **A13. The five least liked candy types are Nik L Nip, Boston Baked Beans, Chiclets, Super Bubble, and Jawbusters.** 

```{r}
# look at the top five lowest ranked candy types
head(candy[order(candy$winpercent),], n=5)
```

> **Q14. What are the top 5 all time favorite candy types out of this set?**

> **A14. The five most liked candy types are Snickers, Kit Kat, Twix, Reese's Miniatures, and Reese's Peanut Butter cup.** 

```{r}
# look at the top five highest ranked candy types
tail(candy[order(candy$winpercent),], n=5)
```

> **Q15. Make a first barplot of candy ranking based on winpercent values.**

```{r}
library(ggplot2)
#create barplot using winpercent values
ggplot(candy) + 
  aes(winpercent, rownames(candy)) +
  geom_col()
```

> **Q16. This is quite ugly, use the reorder() function to get the bars sorted by winpercent?**

```{r}
#reorder based on winpercent (highest to lowest)
ggplot(candy) + 
  aes(winpercent, reorder(rownames(candy),winpercent)) +
  geom_col()
```

```{r}
# add specific color for each type of candy
my_cols=rep("black", nrow(candy))
my_cols[as.logical(candy$chocolate)] = "chocolate"
my_cols[as.logical(candy$bar)] = "brown"
my_cols[as.logical(candy$fruity)] = "pink"
```

```{r}
# create barplot with our color-coded candies
ggplot(candy) + 
  aes(winpercent, reorder(rownames(candy),winpercent)) +
  geom_col(fill=my_cols) 
```

> **Q17. What is the worst ranked chocolate candy?**

> **A17. The worst ranked chocolate candy is Sixlets.** 

> **Q18. What is the best ranked fruity candy?**

> **A18. The best ranked fruity candy is Starburst.**

# 5. Taking a look at pricepercent

```{r}
# load ggrepel
library(ggrepel)

# How about a plot of win vs price
ggplot(candy) +
  aes(winpercent, pricepercent, label=rownames(candy)) +
  geom_point(col=my_cols) + 
  geom_text_repel(col=my_cols, size=3.3, max.overlaps = 5)
```

> **Q19. Which candy type is the highest ranked in terms of winpercent for the least money - i.e. offers the most bang for your buck?**

> **A19. The highest ranked and cheapest is Reese's Miniatures.**

> **Q20. What are the top 5 most expensive candy types in the dataset and of these which is the least popular?**

> **A20. The five most expensive candy types are Nik L Nip, Nestle Smarties, Ring pop, Hershey's Krackel, and Hershey's Milk Chocolate. Of these, the least popular is Nik L Nip.** 

```{r}
ord <- order(candy$pricepercent, decreasing = TRUE)
head( candy[ord,c(11,12)], n=5 )
```

# 6. Exploring the correlation structure

```{r}
# load corrplot
library(corrplot)
```

```{r}
# create a correlation matrix of the candy dataset
cij <- cor(candy)
corrplot(cij)
```

> **Q22. Examining this plot what two variables are anti-correlated (i.e. have minus values)?**

> **A22. Chocolate and fruity are the most anti-correlated.**

> **Q23. Similarly, what two variables are most positively correlated?**

> **A23. Chocolate and bars are the most positively correlated.**

# 7. Principal Component Analysis

```{r}
# apply PCA to our candy dataset
pca <- prcomp(candy, scale=TRUE)
summary(pca)
```

```{r}
# plot main PCA score plot of PC1 vs. PC2
plot(pca$x[, 1:2])
```

```{r}
# change plotting character and add color
plot(pca$x[,1:2], col=my_cols, pch=16)
```

```{r}
# Make a new data-frame with our PCA results and candy data
my_data <- cbind(candy, pca$x[,1:3])
```

```{r}
# change plot variables
p <- ggplot(my_data) + 
        aes(x=PC1, y=PC2, 
            size=winpercent/100,  
            text=rownames(my_data),
            label=rownames(my_data)) +
        geom_point(col=my_cols)

p
```

```{r}
# load ggrepel
library(ggrepel)

# label the plot, add title and subtitle
p + geom_text_repel(size=3.3, col=my_cols, max.overlaps = 7)  + 
  theme(legend.position = "none") +
  labs(title="Halloween Candy PCA Space",
       subtitle="Colored by type: chocolate bar (dark brown), chocolate other (light brown), fruity (red), other (black)",
       caption="Data from 538")
```

```{r}
# generate interactive plot to see labels
library(plotly)
ggplotly(p)
```

```{r}
# create PCA loadings plot
ggplot(pca$rotation) +
  aes(PC1, reorder(rownames(pca$rotation), PC1)) +
  geom_col() 
```

> **Q24. Complete the code to generate the loadings plot above. What original variables are picked up strongly by PC1 in the positive direction? Do these make sense to you? Where did you see this relationship highlighted previously?**

> **A24. The variables picked up strongly in the positive direction are fruity, pluribus, and hard. The variables picked up strongly in the negative direction are chocolate and bar. These two types of variables have a very high negative correlation when put together, as we can see in the corrplot above. When determining the relationships between chocolate and the three positive variables, they are all orange/red. This means that PC1 is showing the extreme differences in the beginning/end of the plot, fruity and chocolate, and then gradaully shows more positive correlations, and subsequently other types of candy, towards the middle.**

# 8. Summary

> **Q25. Based on your exploratory analysis, correlation findings, and PCA results, what combination of characteristics appears to make a “winning” candy? How do these different analyses (visualization, correlation, PCA) support or complement each other in reaching this conclusion?**

> **A25. Chocolate + bar make a "winning" candy, as does fruity + pluribus + hard. Both chocolate and bar show up together at the end of the PCA barplot, while fruity, pluribus, and hard show up together at the beginning of the barplot, therefore indicating that these characteristics are often grouped together. The corrplot can also attribute to this, as the chocolate + bar combination have the the highest positive correlation, while the fruity + hard combination also also pretty high. Both types of the analyses show similar ideas, with the corrplot also visualizing the intensity of the correlations, and the PCA barplot focusing on the variance.** 