---
title: "class12"
author: "Alishba Qadri"
format: pdf
editor: visual
---

# 3. Import countData and colData

```{r}
# Complete the missing code
counts <- read.csv("airway_scaledcounts.csv", row.names=1)
metadata <-  read.csv("airway_metadata.csv")
```

```{r}
# look at first few rows of dataframe
head(counts)
```

```{r}
# look at first few rows of dataframe
head(metadata)
```

> **Q1. How many genes are in this dataset?**

> **A1. There are 38694 genes in this dataset.**

```{r}
# find number of genes
nrow(counts)
```

> **Q2. How many ‘control’ cell lines do we have?**

> **A2. There are 4 'control' cell lines.**

```{r}
# view entire object, find number of 'control' lines
View(metadata)
```

# 4. Toy differential gene expression

```{r}
# find the sample id for the controls
control <- metadata[metadata[,"dex"]=="control",]
control.counts <- counts[ ,control$id]
# calculate the mean counts per gene across these samples
control.mean <- rowSums( control.counts )/4 
head(control.mean)
```

```{r}
# alternate way to find mean using dplyr package
library(dplyr)
control <- metadata %>% filter(dex=="control")
control.counts <- counts %>% select(control$id) 
control.mean <- rowSums(control.counts)/4
head(control.mean)
```

> **Q3. How would you make the above code in either approach more robust? Is there a function that could help here?**

> **A3. If I add more samples, the values obtained with the exact code above would not be correct, as the code is always going to be dividing by 4, which is assumed to be the number of samples in my data, to get the mean. Therefore, in order to be able to put 'n' amount of samples, I would use the rowMeans() function instead.**

```{r}
# use rowMeans() function to find control sample means
control <- metadata %>% filter(dex=="control")
control.counts <- counts %>% select(control$id) 
control.mean <- rowMeans(control.counts)
head(control.mean)
```

> **Q4. Follow the same procedure for the treated samples (i.e. calculate the mean per gene across drug treated samples and assign to a labeled vector called treated.mean)**

> **A4. 

```{r}
# Identify treated samples
treated <- metadata[metadata[,"dex"]=="treated",]
# calculate the mean counts per gene across these samples 
treated.mean <- rowMeans( counts[ ,treated$id] )
head(treated.mean)
```

```{r}
# combine our meancount data for bookkeeping purposes
meancounts <- data.frame(control.mean, treated.mean)
```

```{r}
# show sum of mean counts across all genes for each group
colSums(meancounts)
```

> **Q5. 

a. Create a scatter plot showing the mean of the treated samples against the mean of the control samples.

```{r}
# create scatterplot
plot(meancounts[,1],meancounts[,2], xlab="Control", ylab="Treated")
```

b. You could also use the ggplot2 package to make this figure producing the plot below. What geom_?() function would you use for this plot? 

> **I would use the geom_point() function. 

> **Q6. Try plotting both axes on a log scale. What is the argument to plot() that allows you to do this?**

> **A6. scale_x_continuous(trans="log2") and scale_y_continuous(trans="log2") allows me to to plot the data using a log2 scale.**

```{r}
# load ggplot2
library(ggplot2)
# create scatteplot using log scale
ggplot(meancounts) +
  aes(x = control.mean, y = treated.mean) +
  geom_point(shape = 1) +
  scale_x_continuous(trans="log2") +
  scale_y_continuous(trans="log2")
```

```{r}
# add calculated log2foldchange to dataframe
meancounts$log2fc <- log2(meancounts[,"treated.mean"]/meancounts[,"control.mean"])
# view first few rows
head(meancounts)
```

```{r}
# find every zero in the table
zero.vals <- which(meancounts[,1:2]==0, arr.ind=TRUE)
# determine specific rows to remove
to.rm <- unique(zero.vals[,1])
# filter out the rows found in 'to.rm'
mycounts <- meancounts[-to.rm,]
# view first few rows
head(mycounts)
```

> **Q7. What is the purpose of the arr.ind argument in the which() function call above? Why would we then take the first column of the output and need to call the unique() function?**

> **A7. arr.ind=TRUE shows which genes (rows) and samples (columns) have zero counts, so we don't take that entire gene into account. We need to call the unique() function so that we don't count a row twice if there is more than one zero from each sample per row.** 

```{r}
# filter the dataset to be expressed as a log2(FoldChange) of greater than 2 or less than -2
up.ind <- mycounts$log2fc > 2
down.ind <- mycounts$log2fc < (-2)
```

> **Q8. Using the up.ind vector above can you determine how many up regulated genes we have at the greater than 2 fc level?**

> **A8. There are 250 up-regulated genes greater than 2 fc level.**

```{r}
# find number of 'up'-regulated genes.
sum(up.ind, na.rm=TRUE)
```

> **Q9. Using the down.ind vector above can you determine how many down regulated genes we have at the greater than 2 fc level?**

> **A9. There are 367 down-regulated genes greater than 2 fc level.**

```{r}
# find number of 'down'-regulated genes. 
sum(down.ind, na.rm=TRUE)
```

> **Q10. Do you trust these results? Why or why not?**

> **A10. I don't trust these results, since we haven't performed any tests to determine whether these differences are actually statistically significant.**

# 5. Setting up for DESeq

```{r}
# load package
library(DESeq2)
citation("DESeq2")
```

```{r}
# create dds dataset 
dds <- DESeqDataSetFromMatrix(countData=counts, 
                              colData=metadata, 
                              design=~dex)
# run analysis
dds <- DESeq(dds)
# view dataset
dds
```

# 6. Principal Component Analysis (PCA)

```{r}
# apply a variance stabilizing transformation
vsd <- vst(dds, blind = FALSE)
# calculate PCs and plot the results
plotPCA(vsd, intgroup = c("dex"))
```

```{r}
# store data into pcaData dataframe
pcaData <- plotPCA(vsd, intgroup=c("dex"), returnData=TRUE)
# view first few rows of dataframe
head(pcaData)
```

```{r}
# Calculate percent variance per PC for the plot axis labels
percentVar <- round(100 * attr(pcaData, "percentVar"))
```

```{r}
# plot pcaData
ggplot(pcaData) +
  aes(x = PC1, y = PC2, color = dex) +
  geom_point(size =3) +
  xlab(paste0("PC1: ", percentVar[1], "% variance")) +
  ylab(paste0("PC2: ", percentVar[2], "% variance")) +
  coord_fixed() +
  theme_bw()
```

# 7. DESeq analysis

```{r}
# access dataset results
results(dds)
```

```{r}
# reassign entire DESeq dataset back to dds
dds <- DESeq(dds)
```

```{r}
# obtain dataset results
res <- results(dds)
res
```

```{r}
# summarize the basics
summary(res)
```

```{r}
# run results on a stricter p-value threshold (0.05)
res05 <- results(dds, alpha=0.05)
summary(res05)
```

# 8, Adding annotation data

```{r}
# load packages
library("AnnotationDbi")
library("org.Hs.eg.db")
```

```{r}
# obtain a list of all avilable key types we can translate
columns(org.Hs.eg.db)
```

```{r}
# create new column in results table
res$symbol <- mapIds(org.Hs.eg.db,
                     keys=row.names(res), # Our genenames
                     keytype="ENSEMBL",        # The format of our genenames
                     column="SYMBOL",          # The new format we want to add
                     multiVals="first")
```

```{r}
# view first few rows of results table
head(res)
```

> **Q11. Run the mapIds() function two more times to add the Entrez ID and UniProt accession and GENENAME as new columns called res$entrez, res$uniprot and res$genename.**

> **A11. 

```{r}
# create Entrez ID column
res$entrez <- mapIds(org.Hs.eg.db,
                     keys=row.names(res),
                     column="ENTREZID",
                     keytype="ENSEMBL",
                     multiVals="first")
# create UniProt accession column
res$uniprot <- mapIds(org.Hs.eg.db,
                     keys=row.names(res),
                     column="UNIPROT",
                     keytype="ENSEMBL",
                     multiVals="first")
# create GENENAME column
res$genename <- mapIds(org.Hs.eg.db,
                     keys=row.names(res),
                     column="GENENAME",
                     keytype="ENSEMBL",
                     multiVals="first")
# view first few rows
head(res)
```

```{r}
# arrange results by adjusted p-value
ord <- order( res$padj )
#View(res[ord,])
head(res[ord,])
```

```{r}
# write out the ordered significant results with annotations
write.csv(res[ord,], "deseq_results.csv")
```

# 9. Data Visualization

```{r}
# visualize data using a Volcano plot
plot( res$log2FoldChange,  -log(res$padj), 
      xlab="Log2(FoldChange)",
      ylab="-Log(P-value)")
```

```{r}
# add some new guidelines
plot( res$log2FoldChange,  -log(res$padj), 
 ylab="-Log(P-value)", xlab="Log2(FoldChange)")

# Add some cut-off lines
abline(v=c(-2,2), col="darkgray", lty=2)
abline(h=-log(0.05), col="darkgray", lty=2)
```

```{r}
# Setup our custom point color vector 
mycols <- rep("gray", nrow(res))
mycols[ abs(res$log2FoldChange) > 2 ]  <- "red" 

inds <- (res$padj < 0.01) & (abs(res$log2FoldChange) > 2 )
mycols[ inds ] <- "blue"

# Volcano plot with custom colors 
plot( res$log2FoldChange,  -log(res$padj), 
 col=mycols, ylab="-Log(P-value)", xlab="Log2(FoldChange)" )

# Cut-off lines
abline(v=c(-2,2), col="gray", lty=2)
abline(h=-log(0.1), col="gray", lty=2)
```

```{r}
# load package
library(EnhancedVolcano)
```

```{r}
# create readable format for the plotting function
x <- as.data.frame(res)
# create volcano plot
EnhancedVolcano(x,
    lab = x$symbol,
    x = 'log2FoldChange',
    y = 'pvalue')
```

