---
title: "Class05"
author: "Alishba"
format: pdf
editor: visual
---

# TITLE HERE

## Subtitle

### Subsubtitle

**stuff in bold**

not in bold

> indent text

*something italics*

> **Q1: Make a ggplot plot**

> **Answer Q1: gherferfer**

```{r}
x <- 1:5
x + 1
```

Let's make a graph

```{r}
library()
```

Fixing Bad code:

```{r}
4 + 2
```

Plots:

```{r}
# loaded ggplot2 package
library(ggplot2)

# specified data set,then determined the aesthetics
ggplot(mpg) +
  aes(x=displ, y=hwy) +
  # visualized data using scatterplot; included trendline
  geom_point() +
  theme_bw() +
  labs(title="Cars", subtitle= "hwy mileage", caption="my data", x="Displacement(L)", y="MPG (Highway)") +
  facet_wrap(~cyl)
```


```{r}
# made another graph using different data set, aesthetics, etc. 
library(ggplot2)
ggplot(cars) +
  aes(x=speed, y=dist) +
  geom_point() +
  labs(title="Speed and Stopping Distances of Cars", x="Speed (MPH)", y="Stopping Distance (ft)", subtitle="Your informative subtitle text here", caption="Dataset: 'cars'") + 
  geom_smooth(method="lm", se=FALSE) +
  theme_bw()
```

> **Q: Which geometric layer should be used to create scatter plots in ggplot2?**

> **A: geom_point()**

In another data set, I created a plot answered questions about gene expression.

```{r}
# created a new data set on anti-viral drugs
url <- "https://bioboot.github.io/bimm143_S20/class-material/up_down_expression.txt"
genes <- read.delim(url)
head(genes)

# found number of genes in data set
nrow(genes)

#found column names and number of columns
colnames(genes)
ncol(genes)

# determined number of 'up' regulated genes
table(genes$State)

#determined fraction of up-regulated genes, rounded it
round( table(genes$State)/nrow(genes) * 100, 2 )

# created basic scatter plot of data set
p <- ggplot(genes) +
  aes(x=Condition1, y=Condition2, col=State) +
  geom_point()
p + scale_colour_manual( values=c("blue", "gray", "red")) +
  labs(title="Gene Expression Changes Upon Drug Treatment", x="Control (no drug)", y="Drug Treatment")
```

> **Q: Use the nrow() function to find out how many genes are in this dataset. What is your answer?**

> **Answer: 5196 genes.**

> **Q: Use the colnames() function and the ncol() function on the genes data frame to find out what the column names are (we will need these later) and how many columns there are. How many columns did you find?**

> **Answer: 4 columns.**

> **Q: Use the table() function on the State column of this data.frame to find out how many ‘up’ regulated genes there are. What is your answer?**

> **Answer: 127 'up' regulated genes.**

> **Using your values above and 2 significant figures. What fraction of total genes is up-regulated in this dataset?**

> **Answer: 2.44% of total genes are up-regulated.**

## Optional Sections 7-10

### Section 7

In this data set, I modeled the economic and demographic data of various countries.

```{r}
# obtained data set
library(gapminder)

# install.packages("dplyr")  ## un-comment to install if needed
library(dplyr)

gapminder_2007 <- gapminder %>% filter(year==2007)

# created plot of the data set for year 2007
ggplot(gapminder_2007) +
  aes(x=gdpPercap, y=lifeExp) +
  geom_point()
```
```{r}
# continuation from previous chunk. made points slightly transparent for better visualization

ggplot(gapminder_2007) +
  aes(x=gdpPercap, y=lifeExp) +
  geom_point(alpha=0.5)
```
```{r}
# continuation from previous chunk. added more variables for better visualization.

ggplot(gapminder_2007) +
  aes(x=gdpPercap, y=lifeExp, color=continent, size=pop) +
  geom_point(alpha=0.5)
```
```{r}
# colored points by numeric variable population 
ggplot(gapminder_2007) + 
  aes(x = gdpPercap, y = lifeExp, color = pop) +
  geom_point(alpha=0.8)
```
```{r}
# continuation of gapminder_2007 data set. plotted GDP per capita vs. life expectancy. adjusted point size to correlate with population size

ggplot(gapminder_2007) + 
  aes(x = gdpPercap, y = lifeExp, size = pop) +
  geom_point(alpha=0.5)
```
```{r}
# made point sizes more proportional to population size

ggplot(gapminder_2007) + 
  geom_point(aes(x = gdpPercap, y = lifeExp,
                 size = pop), alpha=0.5) + 
  scale_size_area(max_size = 10)
```

The following plots are for the year 1957 instead of 2007. 

```{r}
# filtered data set to only include the year 1957
gapminder_1957 <- gapminder %>% filter(year==1957)

# created scatter plot of GDP per capita vs. life expectancy. adjusted variables for better visualization
ggplot(gapminder_1957) +
  aes(x = gdpPercap, y=lifeExp, color=continent, size=pop) +
  geom_point(alpha=0.7) + 
  scale_size_area(max_size=10)
```

```{r}
# completed plots for both years 1957 and 2007
gapminder_1957 <- gapminder %>% filter(year==1957 | year==2007)

ggplot(gapminder_1957) + 
  geom_point(aes(x = gdpPercap, y = lifeExp, color=continent,
                 size = pop), alpha=0.7) + 
  scale_size_area(max_size = 10) +
  # used facet_wrap() to create two plots
  facet_wrap(~year)
```

### Section 8

```{r}
# loaded data set showing people in the five biggest countries by population in 2007
gapminder_top5 <- gapminder %>% 
  filter(year==2007) %>% 
  arrange(desc(pop)) %>% 
  top_n(5, pop)

# created bar chart for country vs. population size
gapminder_top5

ggplot(gapminder_top5) + 
  geom_col(aes(x = country, y = pop))
```

```{r}
# created bar chart showing life expectancy of five most populous countries in 2007

ggplot(gapminder_top5) + 
  geom_col(aes(x = country, y = lifeExp))
```
```{r}
# colored the bars of the country vs. population graph

ggplot(gapminder_top5) + 
  geom_col(aes(x = country, y = pop, fill = continent))
```
```{r}
# colored the bars using numeric variable of life expectancy

ggplot(gapminder_top5) + 
  geom_col(aes(x = country, y = pop, fill = lifeExp))
```
```{r}
# created bar chart of country vs. population. used GDP per capita to color bars

ggplot(gapminder_top5) +
  aes(x=country, y=pop, fill=gdpPercap) +
  geom_col()
```
```{r}
# changed order of bars

ggplot(gapminder_top5) +
  aes(x=reorder(country, -pop), y=pop, fill=gdpPercap) +
  geom_col()
```

```{r}
# filled by country

ggplot(gapminder_top5) +
  aes(x=reorder(country, -pop), y=pop, fill=country) +
  geom_col(col="gray30") +
  guides(fill="none")
```

```{r}
# loaded another data set for arrest data in the US

head(USArrests)
```

```{r}
# plotted flipped bar chart for data set

USArrests$State <- rownames(USArrests)
ggplot(USArrests) +
  aes(x=reorder(State,Murder), y=Murder) +
  geom_col() +
  coord_flip()
```

```{r}
# visualized graph better

ggplot(USArrests) +
  aes(x=reorder(State,Murder), y=Murder) +
  geom_point() +
  geom_segment(aes(x=State, 
                   xend=State, 
                   y=0, 
                   yend=Murder), color="blue") +
  coord_flip()
```

### Section 9

```{r}
# loaded package
library(gapminder)
library(gganimate)

# Setup nice regular ggplot of the gapminder data
ggplot(gapminder, aes(gdpPercap, lifeExp, size = pop, colour = country)) +
  geom_point(alpha = 0.7, show.legend = FALSE) +
  scale_colour_manual(values = country_colors) +
  scale_size(range = c(2, 12)) +
  scale_x_log10() +
  # Facet by continent
  facet_wrap(~continent) +
  # Here comes the gganimate specific bits
  labs(title = 'Year: {frame_time}', x = 'GDP per capita', y = 'life expectancy') +
  transition_time(year) +
  shadow_wake(wake_length = 0.1, alpha = FALSE)
```

### Section 10

```{r}
library(patchwork)

# Setup some example plots 
p1 <- ggplot(mtcars) + geom_point(aes(mpg, disp))
p2 <- ggplot(mtcars) + geom_boxplot(aes(gear, disp, group = gear))
p3 <- ggplot(mtcars) + geom_smooth(aes(disp, qsec))
p4 <- ggplot(mtcars) + geom_bar(aes(carb))

# Use patchwork to combine them here:
(p1 | p2 | p3) /
      p4
```
