---
title: "class07"
author: Alishba Qadri
format: pdf
editor: visual
---

```{r}
# Generate some example data for clustering
tmp_x <- c(rnorm(15, -2), rnorm(15, 2))
tmp_y <- c(rnorm(15, 2), rnorm(15, -2))

df <- cbind(x=tmp_x, y=tmp_y)
```

Plot the data
```{r}
plot(df)
```

Let's do a k-means clustering analysis
```{r}
km <- kmeans(df, centers=2, nstart=20)

km
```

**Q: Number of points per cluster?**
```{r}
km$size
```

Q: Cluster size?
km$size

Q: Cluster assignment
```{r}
km$cluster
```
```{r}
table(km$cluster)
```

Q: Cluster centers
```{r}
km$centers
```

Plot k-means clustering:
```{r}
plot(df, col=c("RED", "BLACK")[km$cluster])
points(km$centers, col="BLUE", pch="X")
```

Plot with ggplot:
```{r}
library(ggplot2)

ggplot(df) +
  aes(x, y, col=factor(km$cluster)) +
  geom_point() +
  scale_color_manual(values=c("RED", "BLACK")) +
  geom_point(data=km$centers, col="BLUE", shape="X", size=3, aes(x, y))

```

# Hierarchical clustering

Analyze by hclust
```{r}
dm <- dist(df)

#dm
```

```{r}
hc <- hclust(dm)

hc
```

```{r}
plot(hc)
abline(h=6, col="RED")
```

Let's use cutree:
```{r}
cutree(hc, h=6)
```

Cut by height
```{r}
table(cutree(hc, h=6))
```

Cut by k
```{r}
cutree(hc, k=2)
```

4 clusters
```{r}
cutree(hc, k=4)
```

```{r}
plot(df, col=cutree(hc, k=2))
```

# PCA of UK food data

## Data import

```{r}
url <- "https://tinyurl.com/UK-foods"
x <- read.csv(url)
```

> **Q1. How many rows and columns are in your new data frame named x? What R functions could you use to answer this questions?**

> **A1: 17 rows and 5 columns.** 
```{r}
# determined number of rows and columns
dim(x)
```

## Checking your data

```{r}
# Previewed the first 6 rows.
head(x)
```

```{r}
# Note how the minus indexing works
rownames(x) <- x[,1]
x <- x[,-1]
head(x)
```

```{r}
#Let's check the dimensions again.
dim(x)
```

```{r}
# testing alternative approach to setting the correct row names
x <- read.csv(url, row.names=1)
head(x)
```

> **Q2: Which approach to solving the ‘row-names problem’ mentioned above do you prefer and why? Is one approach more robust than another under certain circumstances?**

> **A2: I prefer the second approach more, because the first one caused a column to be deleted each time I ran that line of code. The second approach does not have that restriction.  

```{r}
x <- x[,-1]
```
```{r}
x <- read.csv(url, row.names=1)
head(x)
```

## Spotting major differences and trends

```{r}
# Using base R
barplot(as.matrix(x), beside=T, col=rainbow(nrow(x)))
```

> **Q3: Changing what optional argument in the above barplot() function results in the following plot?**

> **A3: Changing the beside argument to F results in stacked bars. T causes juxtaposed bars.**

```{r}
# determining change in plot
barplot(as.matrix(x), beside=F, col=rainbow(nrow(x)))

?barplot
```

```{r}
# Currently we have wide format.
dim(x)
```
```{r}
# data organized with one row per Food and multiple columns for Country
head(x)
```

## Spotting major differences and trends

```{r}
# load tidyr
library(tidyr)

# Convert data to long format for ggplot with `pivot_longer()`
x_long <- x |> 
          tibble::rownames_to_column("Food") |> 
          pivot_longer(cols = -Food, 
                       names_to = "Country", 
                       values_to = "Consumption")

dim(x_long)
```

```{r}
head(x_long)
```

```{r}
# Create grouped bar plot
library(ggplot2)
```

```{r}
# Create bar plot, tweak aesthetics
ggplot(x_long) +
  aes(x = Country, y = Consumption, fill = Food) +
  geom_col(position = "dodge") +
  theme_bw()
```

> **Q4: Changing what optional argument in above ggplot() code results in a stacked barplot figure?**

> **A4: Changing geom_col("dodge") to geom_col("stack") results in a stacked barplot figure.**

```{r}
ggplot(x_long) +
  aes(x = Country, y = Consumption, fill = Food) +
  geom_col(position = "stack") +
  theme_bw()
```

## Pair plots and heatmaps

> **Q5: We can use the pairs() function to generate all pairwise plots for our countries. Can you make sense of the following code and resulting figure? What does it mean if a given point lies on the diagonal for a given plot?**

> **A5: x creates a pairwise matrix. col=raindow(nrow(x)) assigns a color to each food category. pch=16 determines point style. If a given point lies on the diagonal, this means that people from both countries eat a similar amount of food.** 

```{r}
# Generating pairwise plots for our countries
pairs(x, col=rainbow(nrow(x)), pch=16)

```

```{r}
# load pheatmap
library(pheatmap)

# generate heatmap to perform hierarchical clustering
pheatmap( as.matrix(x) )
```

> **Q6. Based on the pairs and heatmap figures, which countries cluster together and what does this suggest about their food consumption patterns? Can you easily tell what the main differences between N. Ireland and the other countries of the UK in terms of this data-set?**

> **A6. England and Wales cluster together. Scotland clusters next, while N. Ireland joins last. This means that England and Wales have the most similar food consumption patterns, then Scotland, and finally N. Ireland. Ireland consumes a higher amount of potatoes, and less fruit and other meat.**

## PCA to the rescue

```{r}
# Use the prcomp() PCA function 
pca <- prcomp ( t(x) )
summary(pca)
```

> **Q7. Complete the code below to generate a plot of PC1 vs PC2. The second line adds text labels over the data points**

```{r}
# Create a data frame for plotting
df <- as.data.frame(pca$x)
df$Country <- rownames(df)

# Plot PC1 vs PC2 with ggplot
ggplot(pca$x) +
  aes(x = PC1, y = PC2, label = rownames(pca$x)) +
  geom_point(size = 3) +
  geom_text(vjust = -0.5) +
  xlim(-270, 500) +
  xlab("PC1") +
  ylab("PC2") +
  theme_bw()
```

> **Q8. Customize your plot so that the colors of the country names match the colors in our UK and Ireland map and table at start of this document.**

```{r}
# define country color vector
country_colors <- c("England" = "orange", "Wales" = "red", "Scotland" = "blue", "N.Ireland" = "darkgreen")

# plot graph
ggplot(pca$x) +
  aes(x = PC1, y = PC2, label = rownames(pca$x)) +
  geom_point(size = 3) +
  geom_text(vjust = -0.5, color = country_colors) +
  xlim(-270, 500) +
  xlab("PC1") +
  ylab("PC2") +
  theme_bw()
```

## Digging deeper (variable loadings)

```{r}
# calculate variation in original data each PC accounts for
v <- round( pca$sdev^2/sum(pca$sdev^2) * 100 )
v
```

```{r}
# alternate method to calculate variation
z <- summary(pca)
z$importance
```

```{r}
# Create scree plot with ggplot
variance_df <- data.frame(
  PC = factor(paste0("PC", 1:length(v)), levels = paste0("PC", 1:length(v))),
  Variance = v
)

ggplot(variance_df) +
  aes(x = PC, y = Variance) +
  geom_col(fill = "steelblue") +
  xlab("Principal Component") +
  ylab("Percent Variation") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 0))
```

```{r}
## Lets focus on PC1 as it accounts for > 90% of variance 
ggplot(pca$rotation) +
  aes(x = PC1, 
      y = reorder(rownames(pca$rotation), PC1)) +
  geom_col(fill = "steelblue") +
  xlab("PC1 Loading Score") +
  ylab("") +
  theme_bw() +
  theme(axis.text.y = element_text(size = 9))
```

> **Q9: Generate a similar ‘loadings plot’ for PC2. What two food groups feature prominantely and what does PC2 maninly tell us about?**

> **A9: Soft drinks and fresh potatoes feature predominantly. PC2 gives us most of the variance not captured by PC1. While PC1 is mainly N.Ireland vs. Great Britain, PC2 is mainly Scotland vs. Wales. 

```{r}
ggplot(pca$rotation) +
  aes(x = PC2, 
      y = reorder(rownames(pca$rotation), PC2)) +
  geom_col(fill = "steelblue") +
  xlab("PC2 Loading Score") +
  ylab("") +
  theme_bw() +
  theme(axis.text.y = element_text(size = 9))
```

# PCR of RNA-seq data (Optional)

```{r}
# load dataset
url2 <- "https://tinyurl.com/expression-CSV"
rna.data <- read.csv(url2, row.names=1)
# generate first six rows
head(rna.data)
```

> **Q10: How many genes and samples are in this data set? How many PCs do you think it will take to have a useful overview of this data set (see below)?**

> **A10: There are 100 genes and 10 samples in the data set. I believe one PC will be sufficient to view this dataset, since PC1 has a variance of 92.62%, which is quite a huge amount.

```{r}
# determine number of genes and samples
dim(rna.data)
```


```{r}
## Again we have to take the transpose of our data 
pca <- prcomp(t(rna.data), scale=TRUE)

# Create data frame for plotting
df <- as.data.frame(pca$x)
df$Sample <- rownames(df)

## Plot with ggplot
ggplot(df) +
  aes(x = PC1, y = PC2, label = Sample) +
  geom_point(size = 3) +
  geom_text(vjust = -0.5, size = 3) +
  xlab("PC1") +
  ylab("PC2") +
  theme_bw()
```

```{r}
# determine the amount of variation each PC accounts for
summary(pca)
```

```{r}
# Calculate variance explained
pca.var <- pca$sdev^2
pca.var.per <- round(pca.var/sum(pca.var)*100, 1)

# Create scree plot data
scree_df <- data.frame(
  PC = factor(paste0("PC", 1:10), levels = paste0("PC", 1:10)),
  Variance = pca.var[1:10]
)

ggplot(scree_df) +
  aes(x = PC, y = Variance) +
  geom_col(fill = "steelblue") +
  ggtitle("Quick scree plot") +
  xlab("Principal Component") +
  ylab("Variance") +
  theme_bw()
```

```{r}
# Percent variance is often more informative to look at 
pca.var.per
```

```{r}
# Create percent variance scree plot
scree_pct_df <- data.frame(
  PC = factor(paste0("PC", 1:10), levels = paste0("PC", 1:10)),
  PercentVariation = pca.var.per[1:10]
)

ggplot(scree_pct_df) +
  aes(x = PC, y = PercentVariation) +
  geom_col(fill = "steelblue") +
  ggtitle("Scree Plot") +
  xlab("Principal Component") +
  ylab("Percent Variation") +
  theme_bw()
```

```{r}
# A vector of colors for wt and ko samples
colvec <- colnames(rna.data)
colvec[grep("wt", colvec)] <- "red"
colvec[grep("ko", colvec)] <- "blue"

# Add condition to data frame
df$condition <- substr(df$Sample, 1, 2)
df$color <- colvec

ggplot(df) +
  aes(x = PC1, y = PC2, color = color, label = Sample) +
  geom_point(size = 3) +
  geom_text(vjust = -0.5, hjust = 0.5, show.legend = FALSE) +
  scale_color_identity() +
  xlab(paste0("PC1 (", pca.var.per[1], "%)")) +
  ylab(paste0("PC2 (", pca.var.per[2], "%)")) +
  theme_bw()
```

```{r}
loading_scores <- pca$rotation[,1]

## Find the top 10 measurements (genes) that contribute
## most to PC1 in either direction (+ or -)
gene_scores <- abs(loading_scores) 
gene_score_ranked <- sort(gene_scores, decreasing=TRUE)

## show the names of the top 10 genes
top_10_genes <- names(gene_score_ranked[1:10])
top_10_genes 
```

